Fast Mini-Batch MCMC for Big Data Parameter Optimization

### 2016-05-13 nips_2016.tex
Added appendix with some details on gaussians, will email to john tomorrow around noontime

\PassOptionsToPackage{numbers, compress}{natbib}
\usepackage{nips_2016}

% to compile a camera-ready version, add the [final] option, e.g.:
%\usepackage[final]{nips_2016}

@@

\usepackage{amsmath,amsthm,color,graphicx,verbatim,listings}
\graphicspath{{figures/}}

@@

% The \author macro works with any number of authors. There are two commands used to separate the
% names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the lines. Using \AND
% forces a line break at that point. So, if LaTeX puts 3 of 4 authors names on the first line, and
% the last on the second line, try using \AND instead of \And before the third author name.

\author{
  David S.~Hippocampus\thanks{Use footnote for providing further
    information about author (webpage, alternative
    address)---\emph{not} for acknowledging funding agencies.} \\
  Department of Computer Science\\
  Cranberry-Lemon University\\
  Pittsburgh, PA 15213 \\
  \texttt{hippo@cs.cranberry-lemon.edu} \\
  %% examples of more authors
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \AND
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
}

@@

and~\cite{RobbinsMonro1951}.

@@

Cutting the Metropolis-Hastings Budget paper \cite{ cutting_mh_2014} proposed that there is a bias-variance trade-off in standard MCMC methods. Bias occurs when sampling from the wrong distribution, however, bias usually decreases fast. Variance occurs when the number of sampling is not enough since we can collect only a limited number of samples in a given amount of time. The recently proposed Stochastic Gradient Langevin Dynamics (SGLD) \cite{langevin_2011} and Stochastic Gradient Fisher Scoring methods are biased because they omit the required Metropolis-Hastings tests.

@@

\subsection{Experiment 1}

\subsection{Experiment 2}

\subsection{Experiment 3}

\section{Conclusion}\label{sec:conclusion}

@@

\section{Section for Appendix Information}

{\color{blue}
Daniel: In case we want to use this.
}

% Daniel: here are example LaTeX codes for figures and tables if we want to use them.

%\begin{figure}[h]
%  \centering
%  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
%  \caption{Sample figure caption.}
%\end{figure}

%\begin{table}[t]
%  \caption{Sample table title}
%  \label{sample-table}
%  \centering
%  \begin{tabular}{lll}
%    \toprule
%    \multicolumn{2}{c}{Part}                   \\
%    \cmidrule{1-2}
%    Name     & Description     & Size ($\mu$m) \\
%    \midrule
%    Dendrite & Input terminal  & $\sim$100     \\
%    Axon     & Output terminal & $\sim$10      \\
%    Soma     & Cell body       & up to $10^6$  \\
%    \bottomrule
%  \end{tabular}
%\end{table}

%\usepackage[pdftex]{graphicx} ...
%\includegraphics[width=0.8\linewidth]{myfile.pdf}

>>>>

\PassOptionsToPackage{numbers, compress}{natbib}
\usepackage{nips_2016}
% to compile a camera-ready version, add the [final] option, e.g.:
%\usepackage[final]{nips_2016}

@@

\usepackage{amsmath,amsthm,color,graphicx,verbatim,listings,enumitem}
\graphicspath{{figures/}}

@@

% The \author macro works with any number of authors. There are two commands used to separate the
% names and addresses of multiple authors: \And and \AND.
% Using \And between authors leaves it to LaTeX to determine where to break the lines. Using \AND
% forces a line break at that point. So, if LaTeX puts 3 of 4 authors names on the first line, and
% the last on the second line, try using \AND instead of \And before the third author name.

%%  David S.~Hippocampus\thanks{Use footnote for providing further
%%    information about author (webpage, alternative
%%    address)---\emph{not} for acknowledging funding agencies.} \\
%%  Department of Computer Science\\
%%  Cranberry-Lemon University\\
%%  Pittsburgh, PA 15213 \\
%%  \texttt{hippo@cs.cranberry-lemon.edu} \\

\author{
  John Canny \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{canny@berkeley.edu}
  \And
  Haoyu Chen \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{haoyuchen@berkeley.edu}
  \And
  Daniel Seita \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{seita@berkeley.edu}
  \And
  Xinlei Pan \\
  Department of Bioengineering \\
  University of California, Berkeley \\
  \texttt{xinleipan@berkeley.edu}
  \And 
  Biye Jiang \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{bjiang@berkeley.edu}
}

@@

We can use stochasticity~\cite{RobbinsMonro1951}.

@@

@@

\subsection{Mixture of Gaussians}\label{ssec:gaussians}

We start with a simple mixture of Gaussians.

\begin{figure}[ht]
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  %\includegraphics[width=0.5\linewidth]{empty}
  \caption{This will be a sequence of images for the Gaussian mixture.}
\end{figure}

{\color{blue}
Daniel: TODO
}

\subsection{Logistic Regression}\label{ssec:logistic}

\begin{figure}[ht]
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  %\includegraphics[width=0.5\linewidth]{empty}
  \caption{This will be a sequence of images for logistic regression.}
\end{figure}

{\color{blue}
Daniel: TODO
}

\subsection{Neural Network Optimization}\label{ssec:nets}

\begin{figure}[ht]
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  %\includegraphics[width=0.5\linewidth]{empty}
  \caption{This will hopefully show results for neural networks on BIDMach.}
\end{figure}

{\color{blue}
Daniel: TODO
}

\section{Conclusions}\label{sec:conclusion}

{\color{blue}
Daniel: TODO
}

@@

\begin{center}
{\Large Outline of Appendix}
\end{center}

In this appendix, we describe the following topics, with an emphasis on clarity and understanding:

\begin{itemize}[noitemsep]
    \item Detailed discussion of the experiment from Section~\ref{ssec:gaussians}.
    \item Detailed discussion of the experiment from Section~\ref{ssec:logistic}.
    \item Detailed discussion of the experiment from Section~\ref{ssec:nets}.
\end{itemize}

\section{Gaussian Mixture Experiment Details}

\subsection{Mathematical Assumptions}

We borrow this example from~\cite{langevin_2011}. Our parameter is a two dimensional vector, denoted
as $\theta = (\theta_1, \theta_2)$, where we have
\begin{equation}
\theta_1 \sim \mathcal{N}(0, \sigma_1^2) \quad \mbox{ and } \quad \theta_2 \sim \mathcal{N}(0, \sigma_2^2)
\end{equation}
where $\mathcal{N}$ indicates the normal distribution (more generally, the multivariate normal). We
consider the above as our prior. Following~\cite{langevin_2011}, we set $\sigma_1^2 = 10$ and
$\sigma_2^2=1$, so the covariance matrix of $\theta$ is $\Sigma = {\rm diag}(10,1)$. Therefore, the
log prior probability we endow on $\theta$ is
\begin{equation}
\log p(\theta) = \log \left(\frac{1}{2\pi\sqrt{10}}\right) - \frac{1}{2}\theta^T\Sigma^{-1}\theta.
\end{equation}
To generate the data, we use the following Gaussian mixture with tied means:
\begin{equation}\label{eq:x_points}
x_i \sim \frac{1}{2}\mathcal{N}(\theta_1, \sigma_x^2) + \frac{1}{2}\mathcal{N}(\theta_1+\theta_2, \sigma_x^2)
\end{equation}
where, again following~\cite{langevin_2011}, we set $\sigma_x^2 = 2$. This means the log likelihood
of a single data instance is
\begin{equation}
\log p(x_i \mid \theta) = \log\left(\frac{1}{4\sqrt{\pi}}\right) +
\log\left(\exp\left(-\frac{1}{4}(x_i - \theta_1)^2\right) + \exp\left(-\frac{1}{4}(x_i -
(\theta_1+\theta_2))^2\right)\right)
\end{equation}
Here is the problem statement. Given some number of i.i.d. data points $x_1, x_2, \ldots, x_N$
generated according to~(\ref{eq:x_points}), determine the posterior distribution\footnote{Note that,
as is typical with Bayesian analysis, we do not concern ourselves with the denominator of the
original (non-log) posterior.} of $\theta$:
\begin{equation}\label{eq:log_post}
\log p(\theta \mid x_1,\ldots,x_n) = \log p(\theta) + \sum_{i=1}^N\log p(x_i \mid \theta).
\end{equation}
Alternatively, if there are too many data points, we may opt to instead pick a point estimate of
$\theta$, generally the MAP estimate. (If $N$ is extremely large, it will cause the posterior to
peak sharply at its modes, reducing distribution estimates to point estimates.) Note that in many
cases, we will need to take a \emph{minibatch estimate} of~(\ref{eq:log_post}). In that case, the
literature generally uses
\begin{equation}\label{eq:scaling_factor}
\log p(\theta \mid x_1,\ldots,x_n) \approx \log p(\theta) + \frac{N}{n} \sum_{i=1}^n\log p(x_i \mid \theta).
\end{equation}
where we only use $n \ll N$ samples, but we must scale up the likelihood contribution by $N/n$. If we
didn't add this scaling factor, then the contribution of the likelihood terms would be weaker. We
we will discuss this scaling factor in more detail later.

Finally, recall earlier that we needed to define $\Delta$. In our context, $\Delta$ is:
\begin{align}
\Delta &= \log \left(\frac{f(\theta') q(\theta_t \mid \theta')}{f(\theta_t) q(\theta'\mid \theta_t)} \right) \\
&= \log p(\theta') - \log p(\theta_t) + \sum_{i=1}^N(\log p(x_i \mid \theta') - \log p(x_i \mid \theta_t)) + \log\left(\frac{q(\theta_t \mid \theta')}{q(\theta' \mid \theta_t)}\right).
\end{align}

\subsection{Experiment Observations}

Figure~\ref{fig:contour} describes a contour plot of the posterior based on the points we have
sampled. The goal is to get distributions that either cover this posterior well (i.e., identifying
that there are two modes), or if the data is too large and we want MAP estimates, we wish to see the
methods stay at one of the modes.

In the following, we only consider random walk proposals, so the term in $\Delta$ containing the
$q$s disappears.  Random walk proposals are bad, but since the proposal is poor, good performance
can only be obtained with a strong MH test, hence why we can use this as a reasonable starting
benchmark.

We use the settings:

\begin{itemize}[noitemsep]
    \item We use $N=10000$ samples of $x_i$.
    \item The covariance of the Gaussian random walk is ${\rm diag}(0.15, 0.15)$.
    \item The mini-batch size is ...
    \item (more settings) ...
    \item For the adaptive sampling method, the tolerance for deciding on a test is ....
    \item The number of iterations is .... One iteration involves one $\theta'$ proposal and a test
    to accept/reject.
\end{itemize}

Figure~\ref{fig:scatter} compares scatter plots of the ``Cutting the MH'' test (hereafter called
``adaptive minibatch'') versus ``Our Method''.

Figure~\ref{fig:cutting_mh} describes the histograms of the adaptive minibatch's batch sizes.

Figure~\ref{fig:diagnostics} describes the histograms of various statistics for our method.

\begin{figure}[ht]
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  %\includegraphics[width=0.5\linewidth]{empty}
  \caption{TODO contour plots}
  \label{fig:contour}
\end{figure}

\begin{figure}[ht]
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  %\includegraphics[width=0.5\linewidth]{empty}
  \caption{Scatter plots}
  \label{fig:scatter}
\end{figure}

\begin{figure}[ht]
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  %\includegraphics[width=0.5\linewidth]{empty}
  \caption{Diagnostic Plots 1 (e.g., histigrams of cutting MH)}
  \label{fig:cutting_mh}
\end{figure}

\begin{figure}[ht]
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  %\includegraphics[width=0.5\linewidth]{empty}
  \caption{Diagnostic Plots 2 (sd, delta, etc)}
  \label{fig:diagnostics}
\end{figure}

{\color{blue}
Daniel: TODO FINISH
}

\section{Logistic Regression Experiment Details}

{\color{blue}
Daniel: TODO
}

\section{Neural Network Experiment Details}

{\color{blue}
Daniel: TODO
}

% Daniel: here are example LaTeX codes for figures and tables if we want to use them.
%\begin{figure}[h]
%  \centering
%  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
%  \caption{Sample figure caption.}
%\end{figure}
%\begin{table}[t]
%  \caption{Sample table title}
%  \label{sample-table}
%  \centering
%  \begin{tabular}{lll}
%    \toprule
%    \multicolumn{2}{c}{Part}                   \\
%    \cmidrule{1-2}
%    Name     & Description     & Size ($\mu$m) \\
%    \midrule
%    Dendrite & Input terminal  & $\sim$100     \\
%    Axon     & Output terminal & $\sim$10      \\
%    Soma     & Cell body       & up to $10^6$  \\
%    \bottomrule
%  \end{tabular}
%\end{table}
%\usepackage[pdftex]{graphicx} ...
%\includegraphics[width=0.8\linewidth]{myfile.pdf}
